---
title: "Mini-Project 01"
author: "Lauren Bautista"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

# Data Visualization Project 01

```{r setup, include=FALSE}
# Libraries
library(tidyverse)
library(plotly) # For interactive visualization
library(viridis) # For colorblind-safe color palettes
library(here) # For consistent file paths
```

## Data Exploration

```{r data-exploration}
# Reading in the data
usbirths <- read_csv(here("data", "us_births_00_14.csv"))

glimpse(usbirths)

# Checking for missing values
sum(is.na(usbirths))

# Summary stats
summary(usbirths)
```
This dataset contains 6 columns with a total of 5,479 observations. There are no missing values. The minimum number of births in a single day is 5728, the maximum is 16081, and the median is 12343. Given that the data is tidy, no cleaning procedures are necessary. 

## Summarize the Data

In order to prepare for the visualizations, three new dataframes are required for analysis of births by day of the week, days in December, and meteorological season. 
```{r data-summary}
# Avg births by day of week
dayweekbirths <- usbirths %>%
  group_by(day_of_week) %>%
  summarize(avg_births = mean(births, na.rm = TRUE))

# Avg births by day in December
daydecbirths <- usbirths %>%
  filter(month == 12) %>%
  group_by(date_of_month) %>%
  summarize(avg_births = mean(births, na.rm = TRUE))

# Total births per meteorological season
seasonbirths <- usbirths %>%
  mutate(
    season = case_when(
      month %in% c(3,4,5) ~ "Spring",
      month %in% c(6,7,8) ~ "Summer",
      month %in% c(9,10,11) ~ "Fall",
      month %in% c(12,1,2) ~ "Winter"
    )
  )
```

