---
title: "Mini-Project 01"
author: "Lauren Bautista"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

# Data Visualization Project 01

```{r setup, include=FALSE}
# Libraries
library(tidyverse)
library(plotly) # For interactive visualization
library(viridis) # For colorblind-safe color palettes
library(here) # For consistent file paths
```

## Data Exploration

```{r data-exploration}
# Reading in the data
usbirths <- read_csv(here("data", "us_births_00_14.csv"))

glimpse(usbirths)

# Checking for missing values
sum(is.na(usbirths))

# Summary stats
summary(usbirths)
```
This dataset contains 6 columns with a total of 5,479 observations. There are no missing values. The minimum number of births in a single day is 5728, the maximum is 16081, and the median is 12343. Given that the data is tidy, no cleaning procedures are necessary. 

## Summarize the Data

In order to prepare for the visualizations, three new dataframes are required for analysis of births by day of the week, days in December, and meteorological season. 
```{r data-summary}
# Avg births by day of week
dayweekbirths <- usbirths %>%
  group_by(day_of_week) %>%
  summarize(avg_births = mean(births, na.rm = TRUE)) %>%
  mutate(day_type = ifelse(day_of_week %in% c("Sat", "Sun"), "Weekend", "Weekday"))

# Avg births by day in Dec 1 - Jan 5 with holidays highlighted
winterholbirths <- usbirths %>%
  filter(month == 12 | (month == 1 & date_of_month <= 5)) %>%
  group_by(month, date_of_month) %>%
  summarize(avg_births = mean(births, na.rm = TRUE), .groups = 'drop') %>%
  mutate(
    avg_births = round(avg_births, 0),
    period_diff = round(avg_births - winter_mean, 0),
    date_label = paste(ifelse(month == 12, "Dec", "Jan"), date_of_month),
    is_holiday = ifelse(
      (month == 12 & date_of_month %in% c(24,25,31)) | (month == 1 & date_of_month == 1), "Holiday", "Regular Day"
    ),
    holiday_name = case_when(
      month == 12 & date_of_month == 24 ~ "Christmas Eve",
      month == 12 & date_of_month == 25 ~ "Christmas Day",
      month == 12 & date_of_month == 31 ~ "New Year's Eve",
      month == 1  & date_of_month == 1  ~ "New Year's Day",
      TRUE ~ "Regular Day"
    )
  )

# Fixing order of Dec to Jan
order_levels <- c(paste("Dec", 1:31), paste("Jan", 1:5))
winterholbirths$date_label <- factor(winterholbirths$date_label, levels = order_levels)

# Total births per meteorological season
seasonbirths <- usbirths %>%
  mutate(
    season = case_when(
      month %in% c(3,4,5) ~ "Spring",
      month %in% c(6,7,8) ~ "Summer",
      month %in% c(9,10,11) ~ "Fall",
      month %in% c(12,1,2) ~ "Winter"
    )
  )
```

## Visualizations

### Visualization 1: Births by Day of the Week
```{r visualization-1, fig.alt="Bar chart showing Tuesday has the highest average births and Sunday has the lowest, with weekends colored differently than weekdays using a viridis palette."}

ggplot(dayweekbirths, aes(x = factor(day_of_week, levels = c("Mon", "Tues", "Wed", "Thurs", "Fri", "Sat", "Sun")), y = avg_births, fill = day_type)) +
  geom_col() +
  geom_text(aes(label = round(avg_births, 0)), vjust = -0.5, size = 3.5) +
  scale_fill_viridis_d(begin = 0.3, end = 0.8) +
  labs(
    title = "Average Births for Each Day of the Week",
    subtitle = "Significant drop in average births during the weekend",
    x = "Day of the Week",
    y = "Average Births",
    fill = "Day Type"
  ) +
  theme_classic() +
  expand_limits(y = 14000)

```
### Visualization 2: December Births (Interactive)
```{r dec-births, fig.width=10, fig.height=6, fig.alt="Interactive bar chart showing a dip in birth rates around Christmas Eve and Christmas Day in December."}

winter_mean <- mean(winterholbirths$avg_births)

winter_plot <- ggplot(winterholbirths, aes(x = date_label, y = avg_births, fill = is_holiday, text = paste0(holiday_name, "<br>Births: ", avg_births, "<br>+/-: ", period_diff))) +
  geom_col() +
  geom_hline(yintercept = winter_mean, linetype = "dashed", color = "black", alpha = 0.7) +
  scale_fill_manual(values = c("Holiday" = "#35608d", "Regular Day" = "grey80")) +
  labs(
    title = "Effect of Holidays on Daily Births",
    x = "Date",
    y = "Average Births",
    fill = "Day Type"
  ) +
  annotate("text", x = 4, y = 10700, label = "Period Average", size = 2.5, textface = "bold")+
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))

ggplotly(winter_plot, tooltip = "text")
```
By utilizing plotly for interactivity, the reader can deeply explore the distinct drops in birth rates between December 1st and January 5th. Hovering over the specific holiday "valleys" reveals not only the exact date and number of births, but also a calculated field showing exactly how far that day deviates from the period average (e.g., Christmas Day showing a drop of 4,605 births). This allows the base chart to maintain a clean, highly accessible two-color design, while still providing the reader with rich, complex summary statistics on demand without cluttering the visual space.

### Visualization 3: Chart Redesign

#### Original Chart
The original version of this chart removed necessary axis labels and used manual colors that are not colorblind friendly. It is also a very simple visualization that may hide underlying information. 

```{r season-births-1, fig.alt="Original box plot of births by season, using non-accessible colors and missing labels."}
ggplot(seasonbirths, aes(x = factor(season, levels = c("Spring", "Summer", "Fall", "Winter")), y = births, fill = season)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Spring" = "darkolivegreen1", 
                               "Summer" = "gold1", 
                               "Fall" = "darkorange", 
                               "Winter" = "cadetblue1")) +
  labs(
    title = "Daily U.S. Births By Season",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```
#### Redesigned Chart
In this redesigned version, I restored the axis labels for clarity. To increase the complexity and better communicate the story of the data, I upgraded the chart to a Violin plot overlaid with a Boxplot. This shows the density and distribution of births much better. Finally, to meet **Accessibility** standards, I applied the `scale_fill_viridis_d()` colorblind-safe palette, and added descriptive `fig.alt` text to the code chunk.

```{r vis-3-after, fig.alt="A redesigned violin and box plot showing summer has the highest density of high-birth days, using a colorblind safe viridis palette."}
ggplot(seasonbirths, aes(x = factor(season, levels = c("Spring", "Summer", "Fall", "Winter")), y = births, fill = season)) +
  geom_violin(alpha = 0.5, trim = FALSE) + 
  geom_boxplot(width = 0.2, color = "black", alpha = 0.8, outlier.shape = NA) +
  scale_fill_viridis_d() + 
  labs(
    title = "Distribution of Daily U.S. Births By Season",
    x = "Season",
    y = "Daily Births"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```
```{r}
seasonbirths %>%
  group_by(season) %>%
  summarize(
    absolute_range = max(births) - min(births),
    middle_50_spread = IQR(births)
  )
```

