---
title: "Mini-Project 02"
author: "Lauren Bautista"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
---

# Data Visualization Project 02

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(plotly)
library(tigris)
library(broom)
library(here)
library(viridis)
library(patchwork)
```

```{r read-data, include=FALSE}
# Reading in the data
summerhits <- read_csv(here("data", "all_billboard_summer_hits.csv")) # Vis 1

floridalakes <- st_read(here("data/Florida_Lakes", "Florida_Lakes.shp")) # Vis 2

housewrox <- read_csv(here("data", "WestRoxbury.csv")) # Vis 3
```

## Data Exploration

### Summer Hits
```{r data-exploration-1}
glimpse(summerhits)

# Checking for missing values
sum(is.na(summerhits))

# Summary
summary(summerhits)
```
The "all_billboard_summer_hits.csv" dataset has 600 rows and 22 columns with zero NA values. The data is tidy, and thus does not need cleaning prior to summarizing. 

### Houses West Roxbury
```{r data-exploration-3}
glimpse(housewrox)

# Checking for missing values
sum(is.na(housewrox)) 

# Summary
summary(housewrox)
```
The "WestRoxbury.csv" dataset has 5802 rows and 14 columns with zero NA values. The data is tidy, and thus does not need cleaning prior to summarizing. 

## Summarize the Data

### Summer Hits
```{r summarize-summerhits}
# Creating decade variable
summerhits <- summerhits %>%
  mutate(decade = (floor(year/10)*10))

# Grouping by decade and finding total songs, average valence, and average energy
summertrends <- summerhits %>%
  group_by(decade) %>%
  summarize(
    total_songs = n(),
    avg_valence = mean(valence, na.rm = TRUE),
    avg_energy = mean(energy, na.rm = TRUE)
  )

summertrends
```
### Houses West Roxbury
```{r summarize-housewrox}
# Removing houses with no build year
housewrox <- housewrox %>%
  filter(`YR BUILT` != 0)

# Grouping by bedrooms to find total number of houses, average total value, average living area, and average lot sqft
housingmodelsummary <- housewrox %>%
  group_by(BEDROOMS) %>%
  summarize(
    total_houses = n(),
    avg_total_value = mean(`TOTAL VALUE`, na.rm = TRUE),
    avg_living_area = mean(`LIVING AREA`, na.rm = TRUE),
    avg_lot_sqft = mean(`LOT SQFT`, na.rm = TRUE)
  )

housingmodelsummary

# Creating BATHROOMS variable to combine full and half baths, changing unit of living area and lot sqft to 100 ft
unithousewrox <- housewrox %>%
  mutate(
    BATHROOMS = `FULL BATH` + 0.5*`HALF BATH`,
    living_area_100 = `LIVING AREA`/100,
    lot_sqft_100 = `LOT SQFT`/100
  )
```
## Visualizations

### Interactive Plot
```{r}
cb_palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7") # Color-blind friendly categorical palette

intplot <- plot_ly(summerhits, x = ~valence, y = ~energy, 
                   size = ~tempo, sizes = c(5,25),
                   color = ~as.factor(decade),
                   colors = cb_palette,
                   text = ~paste("Song:", track_name, "<br>Artist:", artist_name, "<br>Year:", year),
          hoverinfo = "text",
          type = 'scatter', mode = 'markers',
          marker = list(opacity = 0.6, sizemode = 'diameter')) %>%
    layout(title = "Mood vs. Energy in Summer Hits Across Decades",
           xaxis = list(title = "Valence (Happiness)"),
           yaxis = list(title = "Energy"))
intplot

htmlwidgets::saveWidget(intplot,"interactive_plot.html")
```
### Spatial Visualization
```{r include=FALSE}
usstates <- states(cb = TRUE, class = "sf")
```

```{r}
flboundary <- usstates[usstates$NAME == "Florida", ]

ggplot() +
  geom_sf(data = flboundary,
          fill = "grey80",
          color = "grey70",
          size = 0.4) +
  geom_sf(data = floridalakes,
          fill = "#0072B2",
          color = "#0072B2",
          size = 0.2) +
  annotate("text", x = -80.9, y = 26.9, label = "Lake Ocheechobee", size = 3.5, fontface = "italic", color = "#2C3E50") +
  annotate("text", x = -81.8, y = 28.0, label = "High Density:\nCentral FL", 
           size = 3, color = "#2C3E50") +
  annotate("text", x = -86.8, y = 30.4, label = "Low Density:\nPanhandle", 
           size = 3, color = "#2C3E50") +
  labs(
    title = "Spatial Visualization of Florida Lakes",
    subtitle = "Distribution of surface water across the state",
    caption = "Data from Florida Lakes Dataset"
  ) +
  theme_void()
```
```{r}
ggplot() +
  # 1. Plot the base Florida map and Lakes just like before
  geom_sf(data = flboundary, fill = "grey95", color = "grey80", size = 0.4) +
  geom_sf(data = floridalakes, fill = "#56B4E9", color = NA) +
  
  # 2. THE ZOOM LENS: Restrict the coordinates to the Lake Okeechobee area
  coord_sf(
    xlim = c(-81.6, -80.2), # Longitude boundaries (West to East)
    ylim = c(26.4, 27.4),   # Latitude boundaries (South to North)
    expand = FALSE          # Removes extra white space around the edges
  ) +
  
  # 3. Add an annotation right in the center of the lake
  annotate(
    "text", x = -80.9, y = 26.9, 
    label = "Lake Okeechobee", 
    size = 4, fontface = "italic", color = "#2C3E50"
  ) +
  
  # 4. Update the labels to reflect the new regional focus
  labs(
    title = "Surface Water Focus: Lake Okeechobee Basin",
    subtitle = "Highlighting Florida's largest freshwater lake and surrounding watersheds",
    caption = "Data: Florida Lakes Dataset"
  ) +
  
  theme_void() +
  
  # 5. Add a crisp black border to frame the "zoomed" view
  theme(
    panel.border = element_rect(colour = "black", fill = NA, size = 1),
    plot.title = element_text(margin = margin(b = 5)),
    plot.subtitle = element_text(margin = margin(b = 15))
  )
```

